A lipid transfer protein knockout library reveals ORP9-ORP11 dimer mediating PS/PI(4)P exchange at the ER-trans Golgi contact site to promote sphingomyelin synthesis

Version 1

Abstract

Abstract

Numerous lipids are heterogeneously distributed among organelles. Most lipid trafficking between organelles is achieved by a group of lipid transfer proteins (LTPs) that carry lipids using their hydrophobic cavities. The human genome encodes many intracellular LTPs responsible for lipid trafficking and the function of many LTPs in defining cellular lipid levels and distributions is unclear. Here, we created a gene knockout library targeting 90 intracellular LTPs and performed whole-cell lipidomics analysis. This analysis confirmed known and identified new lipid disturbances caused by loss of LTPs. Among these, we found major sphingolipid imbalances in ORP9 and ORP11 knockout cells, two proteins of unknown function in sphingolipid metabolism. ORP9 and ORP11 form a heterodimer to localize at the ER-trans Golgi membrane contact sites, where the dimer exchanges phosphatidylserine (PS) for phosphatidylinositol-4-phosphate (PI(4)P) between the two organelles. Consequently, loss of either protein causes phospholipid imbalances in the Golgi apparatus that results in lowered sphingomyelin synthesis at this organelle. Overall, our LTP knockout library toolbox identifies various proteins in control of cellular lipid levels including the ORP9-ORP11 heterodimer exchanging PS and PI(4)P at the ER-Golgi membrane contact site as a critical step in sphingomyelin synthesis in the Golgi apparatus.

Introduction

Many types of lipids are heterogeneously distributed among various organelles of the cell1. Due to the hydrophobic nature of lipids, their trafficking needs to be facilitated. While vesicular trafficking contributes only to a small portion of lipid trafficking, the majority of lipid trafficking is achieved by a group of lipid transfer proteins (LTPs) that carry lipids using their hydrophobic cavities to stabilize lipids in the aqueous intracellular environment2. LTPs often contain targeting domains, motifs, transmembrane regions, amphipathic helices or surface charge to define their donor and acceptor organelles3. Many LTPs localize at two organelles simultaneously at membrane contact sites, where these organelles come closer to exchange information and material, including lipids4, 5.

The human genome encodes over 150 LTPs; new LTPs and LTP families are being characterized every so often2, 6–13. About 50 of these proteins are secreted and are involved in carrying lipids, metals, lipopolysaccharides, and other small molecules in blood plasma2. The remaining intracellular LTPs are mainly responsible for intracellular lipid trafficking among organelles. Since many lipid-modifying enzymes are localized to different organelles, LTPs also feed metabolic lipid fluxes, eventually defining cellular lipid levels14. For example, LTP-mediated trafficking of ceramide from the endoplasmic reticulum (ER) to the trans Golgi is needed for sphingomyelin synthesis15. While achieving a considerable understanding of some LTPs in the last years, the function of many LTPs in defining cellular levels and distributions remains unclear.

To study LTPs systematically, we designed and created an arrayed gene knockout library targeting 90 intracellular LTPs based on the CRISPR/Cas9 technology. Lipidomics analysis of the library-generated LTP knockout cells confirmed known and identified novel lipid disturbances emerging from loss of LTPs. These included CERT, GLTP, NPC1 and NPC2 knockout cells with altered sphingolipid levels. Furthermore, we identified major sphingolipid imbalances in ORP9 and ORP11 knockout cells, two proteins of unknown function in sphingolipid metabolism. ORP9 and ORP11 form a heterodimer to localize at the ER-trans Golgi membrane contact sites. At this contact site, the ORP9-ORP11 dimer transfers phosphatidylserine (PS) from the ER to the Golgi and phosphatidylinositol-4-phosphate (PI(4)P) in the opposite direction. Consequently, loss of either protein causes phospholipid imbalances in the Golgi apparatus that results in lowered sphingomyelin synthesis capacity at this organelle. Collectively, our LTP knockout library toolbox identified various proteins controlling cellular lipid levels. Among these, we found the ORP9-ORP11 heterodimer defining phospholipid composition of the trans Golgi as a critical step in sphingomyelin synthesis. These findings highlighted how phospholipid and sphingolipid gradients along the secretory pathway are linked at the ER-Golgi membrane contact sites.

Results

A CRISPR knockout library targeting lipid transfer proteins

To understand the function of LTPs in defining cellular levels and distributions, we created an arrayed CRISPR/Cas9 knockout library targeting LTPs. We targeted only the intracellular LTPs in the library and excluded extracellular proteins, such as LIPOCALIN and LBP-BPI-CETP families2. In this library, 90 wells of a 96-well plate were used for targeting LTPs and 6 for non-targeting (NT) controls (Fig. 1A). For optimal gene disruption, we used a lentiviral gene delivery system and 3 guide RNAs per gene – a method also used by others to increase the frequency of gene disruption16. We applied the LTP knockout library to MelJuSo, a human melanoma cell line. The efficiency of the knockout strategy was confirmed by western blotting of selected wells (Fig. 1B).

Lipidomics analysis of a LTP knockout library revealed known and candidate controllers of cellular lipid levels.

(A) Overview of the arrayed gene knockout library targeting LTPs. NT: non-targeting. (B) Western blotting of the selected wells showing the efficiency of the LTP knockout library on MelJuSo cells. gNT: Non-targeting, control cells. (C) Summary plot of the lipidomics analysis: Z-scores calculated for 17 lipid classes are plotted. Each data point represents the mean Z-score of an LTP knockout cell line; size and color of data points represent the standard deviation of Z-scores for 3 experiments on average. PC: phosphatidylcholine, LPC: lyso-phosphatidylcholine, PE: phosphatidylethanolamine, LPE: lyso-phosphatidylethanolamine, PS: phosphatidylserine, PI: phosphatidylinositol, PA: phosphatidic acid, PG: phosphatidylglycerol, DCer: dihydroxyceramide, Cer: ceramide, SM: sphingomyelin, HexCer: hexosylceramide, LacCer: lactosylceramide, DG: diacylglycerol, TG: triacylglycerol, CE: cholesterol ester, FFA: free fatty acids.

Next, we performed whole-cell lipidomics analysis of the LTP knockout cells. To this end, library-generated MelJuSo cells were grown in lipid-depleted serum and analysed using a lipidomics platform detecting 17 different lipid classes thrice on average (Fig. 1A). One of the detected lipid classes, hexosylceramide, can correspond to glucosylceramide and galactosylceramide – two isomeric lipids in mammalian cells. By metabolic chasing of a fluorescent ceramide analogue in cells silenced for glucosylceramide synthase or galactosylceramide synthase; or treated with the glucosylceramide synthase inhibitor PDMP, we validated that glucosylceramide is the major hexosylceramide in MelJuSo cells (Fig. S1). This further substantiated the previous reports of galactosylceramide being present mainly in oligodendrocytes17.

We next calculated the z-scores within each lipid class based on their relative abundance per measurement and plotted their average for LTP knockout cell lines. The analysis revealed many LTP knockout cells with lipid imbalances (Fig. 1C, Fig. S2A). Meanwhile, no z-score above the absolute value of 2 was observed for NT control cells (Fig. S2B). Overall, the lipidomics analysis of the LTP knockout library uncovered numerous candidate regulators of cellular lipid levels.

NPC1 and NPC2 knockout cells accumulate sphingomyelin in lysosomes and the plasma membrane

Lipidomics analysis of the LTP knockout library revealed various anticipated imbalances. Among these candidate regulators of cellular lipid levels, those related to sphingolipid metabolism are notable. This is likely due to many enzymatic steps of sphingolipid metabolism being more linear and lacking the elasticity of phospholipids1, 14. In the analysis, we identified GLTP knockout cells with lowered glucosylceramide levels (Fig. 2A, Fig. S3A-C). Loss of GLTP is recently reported to block ER-to-Golgi anterograde vesicular trafficking that results in lowered glucosylceramide synthesis18.

NPC1 and NPC2 knockout cells accumulate sphingomyelin in lysosomes.

(A) GLTP knockouts show decreased glucosylceramide levels compared to control cells. GlcCer: glucosylceramide (B) NPC1 and NPC2 knockout cells have increased sphingolipids levels compared to control cells. (C) Increased sphingolipid levels in NPC1 and NPC2 knockouts are represented in all sphingolipid classes. (D) Immunofluorescence images of NPC1 and NPC2 knockout cells permeabilized and stained with recombinant GFP-EQT show the accumulation of sphingomyelin in the lysosomes compared to the parental MelJuSo (MJS) cells. Pixel intensities of the images were adjusted evenly. Scale bar 10µm. (E) Flow cytometry analysis of cells stained with EQT-GFP, CTB-Alexa568 and CD63-V450 on the cell surface. Analysis demonstrates the accumulation of sphingomyelin (EQT-GFP) and the glycosphingolipid GM1 (CTB-Alexa568) on the cell surface without affecting the surface protein levels. (F) Mean fluorescence intensities for each staining normalized to control cells from 3 experiments. Red lines and error bars correspond to mean and standard deviations from 3 experiments, respectively. Approximately 2000 cells were analyzed per experiment.

In addition, we observed NPC1 and NPC2 knockout cells with elevated sphingolipid levels (Fig. 1B). Mutations in NPC1 and NPC2 genes are the cause of Niemann-Pick disease, type C that leads to lysosomal accumulation of cholesterol19. Cells lacking NPC1 or NPC2 function also accumulate sphingolipids20. Our observation of elevated sphingolipid levels in NPC1 and NPC2 knockout cells was present in all sphingolipid classes (Fig. 2B, C). Meanwhile phospholipid levels were not altered dramatically in these cells (Fig. S3D). To confirm whether this increase is associated with lysosomal accumulation of sphingolipids, we used a sphingomyelin-binding biosensor based on the equinatoxin secreted by the beadlet anemone Actinia equina (Fig. S3E)21, 22. Immunofluorescence staining of NPC1 and NPC2 knockout cells using the biosensor showed lysosomal accumulation of sphingomyelin (Fig. 2D). Flow cytometry analysis using the sphingomyelin biosensor and the GM1 glycosphingolipid biosensor, the B subunit of cholera toxin, demonstrated that sphingolipid accumulation is also reflected on the cell surface of NPC1 and NPC2 knockouts (Fig. 2E, F). Meanwhile, the surface protein levels were not altered as detected by staining for the tetraspanin protein CD63. Collectively, our results corroborate the sphingolipid accumulation in NPC1 and NPC2 deficient cells and illustrate that the LTP knockout library is a viable tool to study LTPs and their role in regulating cellular lipid levels.

CERT, ORP9 and ORP11 knockout cells demonstrate reduced sphingomyelin levels

Sphingomyelin is the major sphingolipid in mammalian cells, and it is mainly synthesized in the trans Golgi from ceramide delivered from the ER. The non-vesicular trafficking of ceramide from the ER to the trans Golgi is mediated by the ceramide transfer protein CERT15, 23 (Fig. 3B). Our analysis identified CERT knockouts with the lowest sphingomyelin levels (Fig. 1C, 3A). These cells also demonstrate increased glucosylceramide levels, likely as a compensation mechanism (Fig. 3A). Furthermore, we found decreased sphingomyelin and increased glucosylceramide levels in ORP9 and ORP11 knockout cells – two proteins of unknown role in sphingolipid metabolism (Fig. 1C, 3A). The sphingolipid imbalances in all three knockout cells –CERT, ORP9, and ORP11– were also represented in many sphingolipid subspecies (Fig. 3C, Fig. S4B). Lowered sphingomyelin levels of the three knockout cells were further confirmed by their sensitivity towards methyl-β-cyclodextrin treatment (Fig. 3D). Meanwhile, none of the three knockout cells demonstrated phospholipid imbalances to a compelling degree (Fig. S4A).

CERT, ORP9 and ORP11 knockout cells demonstrate reduced sphingomyelin levels.

(A) CERT, ORP9 and ORP11 knockouts demonstrate decreased sphingomyelin and increased glucosylceramide levels. (B) Schematic representation of de novo sphingomyelin synthesis pathway of mammalian cells. (C) Decreased sphingomyelin and increased glucosylceramide levels observed in knockout cells are present in various sphingolipid subspecies. (D) Similar to CERT knockouts, ORP9 and ORP11 knockouts are sensitive to methyl-β-cyclodextrin treatment. Red lines correspond to mean values from 4 experiments. (E) Western blot of knockout cells showing that the loss of ORP9 or ORP11 does not affect CERT protein levels. (F) ORP9 and ORP11 localize at the Golgi apparatus. Scale bar 10µm. (G) Domain architecture of CERT, ORP9 and ORP11. All proteins contain PH domains for Golgi localization. CERT and ORP9, but not ORP11, contain a FFAT motif for interacting with VAP proteins.

The observation that ORP9 and ORP11 knockout cells mimic CERT knockouts implied a role for ORP9 and ORP11 in de novo sphingomyelin synthesis, whereas loss of neither ORP9 nor ORP11 reduced CERT protein levels (Fig. 3E). ORP9 and ORP11 each contain a PH domain mediating localization at the trans Golgi, the site of de novo sphingomyelin synthesis (Fig.3F, G). As loss of either ORP9 or ORP11 did not alter the Golgi ultrastructure as detected by light or electron microscopy (Fig. S5), it further suggested that ORP9 and ORP11 affect sphingomyelin levels by a previously unknown mechanism. Next, we investigated the possible mechanism by which ORP9 and ORP11 regulate cellular sphingolipid levels.

ORP9 and ORP11 dimerization is critical for their localization to the ER-Golgi contact sites

ORP9 and ORP11 interact with each other via their regions between their PH and ORD domains24. As no other domain is found in this region, we speculated some secondary structures must facilitate the dimerization. AlphaFold and PCOILS coiled coils prediction tool suggested that ORP9 and ORP11 contain two alpha helices each in this region – hereafter referred as “coils” (Fig. 4A, S6A, B). AlphaFold-Multimer suggested that the coils of ORP9 and ORP11 interact with each other (Fig. 4B, S6C-D). Identical analysis of the full-length proteins predicted that the coils drive dimerization of the two proteins (Fig. 4C, D, S6E). This was tested by targeting the coils of ORP9 to the mitochondrial outer membrane using the N-terminal sequence of TOM70. In cells expressing mitochondria-targeted ORP9 coils, the coils of ORP11 located to this organelle (Fig. 4C, D). The absence of either coil prevented the colocalization, suggesting that the coils of ORP9 and ORP11 are sufficient for their dimerization. This finding was validated by co-immunoprecipitations (Fig. 4E).

ORP9 dimerization is critical for ER localization of ORP11.

(A) AlphaFold revealing coiled coils in ORP9 and ORP11. (B) AlphaFold-Multimer shows that coiled coils of ORP9 and ORP11 interacting with each other. (C) ColabFold protein complex prediction for full length ORP9 and ORP11 shows the dimerization via coiled coils (arrow). (D) Position aligned error for panel C. Note that the interaction of coils has a lower score (yellow arrows). (E) Coiled coils of ORP9 and ORP11 is sufficient to define their localization as ORP11 coiled coils colocalizes with the mitochondria-targeted ORP9 coiled coils at mitochondria. Scale bar: 10µm. (F) Colocalization analysis of immunofluorescent images corresponding to panel E. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (G) Co-immunoprecipitation confirming that the coils of ORP9 and ORP11 interact with each other. (H) Immunofluorescence images of PH domain localizations. Compared to individual PH domains of ORP9 and ORP11, ORP9-ORP11 chimera demonstrates better affinity towards the Golgi. Scale bar: 10µm. (I) Colocalization analysis of immunofluorescent images corresponding to panel H. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (J) Co-immunoprecipitation analysis of VAPA, ORP9 and ORP11 from over-expressing cells. Despite lacking a FFAT motif, ORP11 interacts with the ER-resident VAPA protein. This interaction is facilitated by the FFAT motif of ORP9 as a VAPA mutant unable to interact with FFAT motifs (VAPA**) was unable to co-precipitate ORP9 or ORP11.

ORP9, but not ORP11, contains a FFAT, two phenylalanines in an acidic tract, motif that drives ER localization by interacting with the ER-resident VAPA and VAPB25. Despite not containing a FFAT motif, ORP11 could interact with VAPA and VAPB by its dimerization with ORP9. We tested this by co-immunoprecipitations, where ORP11 could be co-isolated with VAPA from cells expressing VAPA, ORP9 and ORP11 (Fig. 4J). This interaction was diminished significantly when ORP9 was not co-expressed. Furthermore, a VAPA mutant unable to interact with FFAT motifs failed to interact with either ORP9 or ORP11, showing that ORP11 interacts with VAPA indirectly via the ORP9-FFAT motif.

We and others previously showed that in addition to VAPA and VAPB, the human proteome contains three other proteins interacting with FFAT and related short linear motifs: MOSPD1, MOSPD2 and MOSPD326, 27. In our earlier efforts to identify interaction partners of VAPA, VAPB and other motif-binding proteins, we performed BioID, proximity biotinylation followed by proteomics, identification of membrane contact sites proteins27. In this analysis, ORP9 and ORP11 were found in proximity with VAPA and VAPB but not with other motif-binding proteins (Fig. S6F). This further confirmed that despite lacking a FFAT motif, ORP11 together with ORP9 is part of VAPA- and VAPB-mediated membrane contact sites.

The Golgi localization of ORP9 and ORP11 is mediated by their PH domain interacting with phosphatidylinositol phosphates24. We observed that the PH domains of ORP9 and ORP11 localized only partially to the Golgi, unlike the PH domains of OSBP and CERT that show exclusive localization to this organelle28, 29 (Fig. 4H, S7D). We hypothesized that the dimerization of ORP9 and ORP11 is a mechanism to increase their avidity towards trans Golgi membranes. To test this, we created a chimera containing ORP9- and ORP11-PH domains that localized to the Golgi more efficiently than the individual PH domains (Fig. 4H, I). This also implied that loss of either protein should reduce the other’s localization to the trans Golgi. Immunofluorescence staining of ORP9 and ORP11 knockout cells for the same proteins confirmed that both ORP9 and ORP11 are required for the Golgi localization of the other protein (Fig. 5). Meanwhile, loss of either protein did not influence the levels of the other (Fig. 3E). Our results collectively show that the dimerization of ORP9 and ORP11 via their coils is required for the localization of ORP9-ORP11 dimer to the ER and the trans Golgi simultaneously.

ORP9 and ORP11 dimerization is critical for their Golgi localization.

(A) Immunofluorescence images of cells stained for ORP9. ORP9 fails to localize at the Golgi as efficiently in ORP11 knockout cells. (B) Colocalization analysis of immunofluorescent images corresponding to panel A. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (C) Immunofluorescence images of cells stained using an ORP11 antibody. Similar to ORP9, ORP11 fails to localize at the Golgi as efficiently in ORP9 knockout cells. Note that the effect of ORP9 loss on ORP11 localization is more dramatic than vice versa. (D) Colocalization analysis of immunofluorescent images from panel C. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. All scale bars are 10µm.

ORP9 and ORP11 are essential for PS and PI(4)P levels in the Golgi apparatus

ORP9 and ORP11 belong to the OSBP-related protein (ORP) family that transfers sterols or PS while transferring phosphatidylinositol phosphates in the opposite direction30–33 (Fig. 6A, S7B). ORP9 and ORP11 carry the conserved PS binding site and are recently shown to relocate to the site of lysosomal damage to supply PS for membrane repair34. Using an in vitro lipid transfer assay, we confirmed that the lipid transfer domains (ORD) of ORP9 and ORP11 are sufficient to traffic PS in vitro (Fig. 6B, S7C). Moreover, similar to ORP5 and ORP8, the PS-trafficking activity of ORP9- and ORP11-ORDs was improved when acceptor liposomes were decorated with PI(4)P35, 36 (Fig. 6C). This indicated that ORDs of ORP9 and ORP11 transport PI(4)P in the opposite direction of PS trafficking.

ORP9 and ORP11 are essential for PS and PI(4)P levels in the Golgi apparatus.

(A) Phylogenetic tree of human OSBP-related domains showing that ORP9 and ORP11 belong to the PS transporter branch. (B) Schematic representation of FRET-based lipid transfer assay. Rhodamine in the donor vesicles quenches NBD fluorescence unless NBD-labeled lipid is transferred to acceptor liposomes. (C) PS transfer activity of ORP9-ORD and ORP11-ORD (ORD9 and ORD11, respectively) is promoted by addition of PI(4)P to acceptor liposomes, suggesting counter transfer of PI(4)P. (D) The PI(4)P sensor OSBP-PH localizes more prominently to the Golgi area in ORP9 and ORP11 knockout cells. (E) Golgi quantification of the PI(4)P sensor confirms the accumulation of PI(4)P at this organelle. This phenotype was rescued by reconstitution of the missing protein in knockout cells (shown with a dagger). Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (F) The PS sensor Lact-C2 localizes less prominently to the Golgi in ORP9 and ORP11 knockout cells. (G) Golgi quantification of the PS sensor RFP-LactC2 indicates reduced PS levels in this organelle. This phenotype could be rescued by reconstitution of the missing protein in knockout cells (shown with a dagger). Red lines correspond to mean values from 3 experiments; n is number of analyzed cells.

PS is synthesized in the ER and its concentration increases along the secretory pathway as it is enriched on the cytosolic leaflet of the plasma membrane14. PI(4)P, on the other hand, is abundant in the trans Golgi, where it recruits many proteins to this membrane37. Another LTP localized at the ER-trans Golgi membrane contact sites, OSBP transfers cholesterol from the ER to the trans Golgi, while counter transporting PI(4)P to the ER for its hydrolysis by the ER-resident PI(4)P phosphatase SAC1 (Fig. 8). Homologous to OSBP, the ORP9-ORP11 dimer resides in the ideal intracellular interface to traffic PS in anterograde and PI(4)P in retrograde direction between the ER and trans Golgi (Fig. 8). This also suggested that the loss of either protein would cause PS and PI(4)P imbalances between the ER and trans Golgi. We tested this idea by using PI(4)P- and PS-binding biosensors38, 39 (Fig. S7D). Golgi localization of these sensors indicated PI(4)P accumulation in the Golgi of ORP9 and ORP11 knockout cells (Fig. 6D, E). Also, PS levels were decreased in the Golgi of these cells (Fig. 6F, G). Both phenotypes were pronounced more strongly in ORP9 knockout cells compared to ORP11 knockouts. Furthermore, both phenotypes, lowered PS and increased PI(4)P levels, could be rescued by reconstitution of the missing protein. Overall, these findings suggested that the ORP9-ORP11 dimer is required for maintaining PS and reducing PI(4)P levels in the Golgi apparatus.

ORP9 and ORP11 are needed for de novo sphingomyelin synthesis in the Golgi apparatus

Next, we set out to investigate the possible role of ORP9 and ORP11 in regulating cellular sphingomyelin levels. By performing a sphingomyelin synthase activity assay that uses the fluorescent ceramide analogue NBD-ceramide as a substrate40, we found that ORP9 and ORP11 knockout cells do not have reduced sphingomyelin synthesis capacity (Fig. 7A, B). The same was also found in CERT knockout cells. Whereas all three knockouts demonstrated increased glucosylceramide synthesis capacity, supporting the lipidomics analysis (Fig. 3A). Unaltered sphingomyelin synthesis capacity was further validated by the unreduced protein and mRNA levels of both human sphingomyelin synthases, SMS1 and SMS2 (Fig. S8A, B).

ORP9 and ORP11 are needed for de novo sphingomyelin synthesis in the Golgi apparatus.

(A) Thin-layer chromatography readout of the enzymatic activity assay performed in lysates reveals the unreduced sphingomyelin synthesis capacity of knockout cells. (B) Quantification of the enzymatic activity assays corresponding to panel A. Note the increased GlcCer production capacity in the knockout cells. Red lines and error bars correspond to mean and standard deviation from 5 experiments, respectively. (C) Graphic representation of experiments using palmitic acid alkyne for monitoring de novo sphingomyelin synthesis. Alkyne modified sphingolipids are “clicked” with azido-coumarin before thin-layer chromatography (TLC) analysis. (D) TLC readout of de novo sphingomyelin synthesis assay in intact cells using palmitic acid alkyne. Knockout cells demonstrate reduced conversion of palmitic acid to sphingomyelin. FA: fatty acid, SM: sphingomyelin, LPE: lyso-O-phosphatidylethanolamine (E) Quantification of de novo sphingomyelin synthesis assay from panel C. Red lines and error bars correspond to mean and standard deviation from 5 experiments, respectively. (F) Simplified representation of Golgi-targeted lipidomics and enzymatic activity assays. (G) Western blot confirming the immunomagnetic isolation of trans Golgi membranes. Cross-reactivity to Protein G and/or Golgin-97 IgG were labeled with asterisks. (H-J) Lipidomics analysis of Golgi isolates show lowered PS levels in the Golgi of ORP9 and ORP11 knockout cells. Same Golgi fractions have reduced sphingomyelin but increased ceramide levels. Each colored line corresponds to a lipid species. Data points and the error bars correspond to mean and standard deviations from 4 experiments. (K) TLC readout of enzymatic activity assay performed in isolated Golgi from knockout cells. Golgi of ORP11 and especially that of ORP9 demonstrate lowered capacity to synthesize sphingomyelin. (L) Quantification of sphingomyelin synthesis activity assay performed in Golgi isolates corresponding to panel I.

SMS1 is localized at the trans Golgi, while SMS2 shows dual localization at the plasma membrane but is also found at the trans Golgi41. As the enzymatic activity assay in lysates reports on the global sphingomyelin synthesis capacity, we next investigated de novo sphingomyelin synthesis that occurs at the trans Golgi, where CERT as well as the ORP9-ORP11 dimer localize. For this purpose, we chased the metabolic fate of palmitic acid alkyne (Fig. 7C). Palmitic acid is the precursor of sphinganine and all sphingolipids (Fig. 3B, 7C). To enter the sphingolipid pathway, palmitic acid first needs to travel to the ER and later to the Golgi for conversion to sphingomyelin, unlike ceramide analogues, such as NBD-ceramide, that can be converted to sphingomyelin by the plasma membrane-resident SMS2. Furthermore, the palmitic acid alkyne does not contain a bulky fluorescent group and the terminal alkyne allows visualization by click chemistry (Fig. 7C). As this method enabled us to monitor de novo sphingomyelin production in intact cells, we confirmed the reduced de novo sphingomyelin synthesis in CERT knockout cells using this method (Fig. 7D, E, S8C). In addition, ORP9 and ORP11 knockouts also showed decreased conversion of palmitic acid to sphingomyelin, suggesting a similar defect as CERT knockouts.

The observations that loss of ORP9 or ORP11 does not affect CERT protein (Fig. 3E) or reduce CERT localization to the Golgi (Fig. S9) implied that CERT-mediated transfer routes are not affected in ORP9 or ORP11 knockouts. Lipidomics analysis showing accumulation of ceramide and glucosylceramide in ORP9/ORP11 knockouts further suggested the conversion of ER-bound ceramides to sphingomyelin in the Golgi is hampered.

To distinguish between two possibilities that may result in lowered sphingomyelin levels – a ceramide delivery defect to the trans Golgi or an inability to convert ceramide to sphingomyelin in the Golgi, we established an immunoisolation protocol for trans Golgi membranes using a monoclonal antibody against the trans Golgi marker Golgin-97 (Fig. 7F). These isolates were enriched for the trans Golgi marker Golgin-97 and were devoid of other organelle markers, such as the ER, mitochondria, and lysosomes (Fig. 7G). We confirmed the membrane integrity of Golgin-97-enriched fractions by their sphingomyelin synthase activity, as these multi-pass transmembrane proteins require intact membranes for activity41 (Fig. S10A). Lipidomics analysis of these fractions validated that the loss of ORP11 and especially that of ORP9 results in lowered PS levels in the trans Golgi (Fig. 7H, S10B-D). The trans Golgi fractions also showed reduced sphingomyelin levels, confirming the limited de novo sphingomyelin synthesis (Fig. 7D, E, I S10E). Moreover, the same fractions showed elevated ceramide levels in these fractions (Fig. 7J, S10F), revealing that the knockouts do not have a ceramide delivery defect, but instead a lowered capacity to convert ceramide to sphingomyelin in their trans Golgi membranes. We tested this notion by performing a sphingomyelin synthase activity assay in the isolated Golgi membranes, where supplying excess substrate to these membranes bypasses the ceramide delivery routes (Fig. 7K, L). This assay further substantiated that the Golgi of ORP11 and especially of ORP9 knockout cells have a lowered sphingomyelin synthesis capacity. Meanwhile, localization of both sphingomyelin synthases was not reduced in the Golgi, as detected by three different antibodies (Fig. S11). Collectively, our results indicate that PS-PI(4)P exchange between the ER and trans Golgi mediated by the ORP9-ORP11 LTP complex is critical for de novo sphingomyelin synthesis in trans Golgi.

Discussion

In this study, we describe a gene knockout library for systematic characterization of intracellular LTPs. The arrayed nature of the library enables high and low throughput analysis. As new LTPs are identified in regular basis, the arrayed feature of the library allows expansion to include newly identified LTPs, e.g. ATG2A, SHIP164, KIAA0100 (Hobbit in Drosophila), KIAA1109 (Csf1 in yeast) and RMDN3 are new human LTPs that are identified through the course of this study8–10, 12, 13.

Lipidomics analysis of the library demonstrated many lipid imbalances. We further validated loss of NPC1, NPC2, CERT, and GLTP with sphingolipid imbalances15, 18, 20 (Fig. 1C). In addition to sphingolipids imbalances, we found STARD7 knockout cells with increased levels of phosphatidylglycerol (PG), a lipid class exclusive to mitochondria in mammals (Fig. 1C). STARD7 is a phosphatidylcholine (PC) transfer protein that localizes to the mitochondrial intermembrane space to supply PC to the mitochondrial inner membrane42–44. Consequently, loss of STARD7 reduces PC levels in this membrane as well as decreasing cardiolipin levels and respiratory capacity43, 44. While PC in the inner membrane is not a precursor of cardiolipin, it is possible the reduced PC levels in this membrane impair the activity of the cardiolipin synthase that uses PG as its substrate, suggesting an explanation for the PG accumulation in STARD7 knockout cells.

LTPs often localize at membrane contact sites to facilitate lipid exchanges between organelles. We found ORP9 and ORP11 localizing at the ER-trans Golgi contact sites to exchange PS and PI(4)P. We show that ORP9-ORP11 dimerization is critical for their contact site localization, and consequently, loss of either protein is sufficient to disturb PS and PI(4)P levels in the Golgi. In return, this phospholipid imbalance reduces sphingomyelin synthesis capacity of the Golgi. The contact site localization of ORP9-ORP11 dimer is homologous to that of ORP9-ORP10 dimer localizing at the ER-endosome contact sites. ORP10 also lacks a FFAT motif and interacts with ORP945. In both LTP complexes, ORP9 provides the FFAT motif required for ER contact. Compared to ORP11 knockout cells, ORP9 knockout cells provide more pronounced phenotypes of cellular sphingolipid imbalances (sphingomyelin and glucosylceramide), phospholipid imbalances in the Golgi apparatus (PS and PI(4)P), as well as reduced capacity of sphingomyelin synthesis in trans Golgi membranes. Furthermore, loss of ORP9 led to a more dramatic effect on the Golgi localization of ORP11 than vice versa. This is most likely due to ORP9 providing a critical factor, i.e. the FFAT motif, for contact site localization, whereas it is possible the loss of ORP11 can be partially compensated by ORP10. ORP10 can also deliver PS to Golgi membranes46, however, our lipidomics analysis of ORP10 knockout cells did not demonstrate any sphingolipid imbalances. In response to lysosomal membrane damage, ORP9, ORP10 and ORP11, together with OSBP and ORP1L localize at the lysosomal membrane to provide PS and other lipids for the repair of the damaged membrane. This localization is driven by a burst of PI(4)P generation as a result of membrane damage, highlighting the role of PI(4)P in spatiotemporal control of lipid trafficking and membrane contact site formation34, 47.

An intriguing observation is that the loss of ORP9 causes more accumulation of ceramide than CERT knockouts, despite loss of either protein leading to a comparable reduction of sphingomyelin levels. As the loss of CERT would lead to a ceramide accumulation in the ER, it is possible that such an accumulation in the ER is sensed to reduce ceramide production in this organelle. An ER-resident candidate ceramide sensor for this purpose was suggested previously48. Meanwhile, our findings show a ceramide accumulation in the trans Golgi of ORP9 knockouts. Accordingly, ceramide accumulation in the Golgi caused by the loss of ORP9 would fail to “turn on” an ER-localized ceramide sensing machinery, thus leading to further ceramide accumulation compared to CERT knockout cells. This notion also supports our finding that the loss of the ORP9-ORP11 dimer causes ceramide accumulation primarily in the Golgi.

Asymmetric distribution of lipids between two bilayer leaflets is a characteristic feature of the plasma membrane. This is owed to the build-up of sphingomyelin and cholesterol on the outer leaflet and PS on the inner leaflet14. The transition of the thin, symmetrical ER membrane to a thicker, rigid and asymmetrical one takes place in the Golgi (Fig. 8)14. Sphingolipid, cholesterol and PS concentrations also increase along the secretory pathway. Various mechanisms are described to drive the sphingolipid and cholesterol trafficking against the concentration gradient, including thermodynamic trapping of cholesterol due to complex formation with sphingomyelin or energy release from the retrograde trafficking of PI(4)P followed by its hydrolysis in the ER14, 31, 49. The same PI(4)P gradient between the ER and Golgi could also power the anterograde trafficking of PS, which, unlike sphingomyelin or cholesterol, is still exposed on the cytosolic leaflet of the Golgi where it has a higher concentration than the ER – thus, maintaining lipid trafficking against the gradient requires energy. In brief, our finding that the ORP9-ORP11 dimer mediated phospholipid exchange promoting the sphingomyelin synthesis reveal further intertwining of lipid gradients along the secretory pathway at the ER-trans Golgi membrane contact sites.

Model of ORP9-ORP11 mediated promotion of sphingomyelin synthesis at the ER-trans Golgi contact site.

ORP9 and ORP11 require dimerization with each other to localize at ER-trans Golgi membrane contact sites, where they exchange PS for PI(4)P. Consequently loss of either protein causes PS and PI(4)P imbalances in the Golgi apparatus. Since only ORP9 contains a FFAT motif, its loss leads to a more pronounced effect on PS, PI(4)P and sphingolipid levels. The same contact site also accommodates OSBP and CERT for the anterograde trafficking of cholesterol and ceramide, the latter is used for sphingomyelin production. Cholesterol, sphingomyelin and PS are trafficked to the plasma membrane by vesicular means to be asymmetrically distributed between the leaflets of the plasma membrane.

Conflict of Interest

Authors declare no conflict of interest.

eLife Assessment

This valuable manuscript systematically addresses the role of intracellular lipid transfer proteins on cellular lipid levels. The authors have provided solid evidence on the role of ORP9 and ORP11 in sphingolipid metabolism at the Golgi complex. The data is in general convincing, except for the claim that ORP9/11 might counter-exchange phosphatidylinositol 4-phosphate and phosphatidylserine, which is not fully supported by the data presented. This article will be of broad interest to cell biologists interested in lipid metabolism and membrane biology.

Reviewer #1 (Public Review):

Summary:

In this well-designed study, the authors of the manuscript have analyzed the impact of individually silencing 90 lipid transfer proteins on the overall lipid composition of a specific cell type. They confirmed some of the evidence obtained by their own and other research groups in the past, and additionally, they identified an unreported role for ORP9-ORP11 in sphingomyelin production at the trans-Golgi. As they delved into the nature of this effect, the authors discovered that ORP9 and ORP11 form a dimer through a helical region positioned between their PH and ORD domains.

Strengths:

This well-designed study presents compelling new evidence regarding the role of lipid transfer proteins in controlling lipid metabolism. The discovery of ORP9 and ORP11's involvement in sphingolipid metabolism invites further investigation into the impact of the membrane environment on sphingomyelin synthase activity.

Weaknesses:

There are a couple of weaknesses evident in this manuscript. Firstly, there's a lack of mechanistic understanding regarding the regulatory role of ORP9-11 in sphingomyelin synthase activity. Secondly, the broader role of hetero-dimerization of LTPs at ER-Golgi membrane contact sites is not thoroughly addressed. The emerging theme of LTP dimerization through coiled domains has been reported for proteins such as CERT, OSBP, ORP9, and ORP10. However, the specific ways in which these LTPs hetero and/or homo-dimerize and how this impacts lipid fluxes at ER-Golgi membrane contact sites remain to be fully understood.

Regardless of the unresolved points mentioned above, this manuscript presents a valuable conceptual advancement in the study of the impact of lipid transfer on overall lipid metabolism. Moreover, it encourages further exploration of the interplay among LTP actions across various cellular organelles.

Reviewer #2 (Public Review):

Summary:

The authors set out to determine which lipid transfer proteins impact the lipids of Golgi apparatus, and they identified a reasonable number of "hits" where the lack of one lipid transfer protein affected a particular Golgi lipid or class of lipids. They then carried out something close to a "proof of concept" for one lipid (sphingomyelin) and two closely related lipid transfer proteins (ORP9/ORP11). They looked into that example in great detail and found a previously unknown relationship between the level of phosphatidylserine in the Golgi (presumably trans-Golgi, trans-Golgi Network) and the function of the sphingomyelin synthase enzyme. This was all convincingly done - results support their conclusions - showing that the authors achieved their aims.

Impact:

There are likely to be 2 types of impact:

(I) cell biology: sphoingomyelin synthase, ORP9/11 will be studied in the future in more informed ways to understand (a) the role of different Golgi lipids - this work opens that out and produces more questions than answers (b) the role of different ORPs: what distinguishes ORP11 from its paralogy ORP10?

(ii) molecular biochemistry: combining knockdown miniscreen with organelle lipidomics must be time-consuming, but here it is shown to be quite a powerful way to discover new aspects of lipid-based regulation of protein function. This will be useful to others as an example, and if this kind of workflow could be automated, then the possible power of the method could be widely applied.

Strengths:

Nicely controlled data;

Wide-ranging lipidomics dataset with repeats and SDs - all data easily viewed.

Simple take-home message that PS traffic to the TGN by ORP9/11 is required for some aspect of SMS1 function.

Weaknesses:

Model and Discussion:

No idea about the aspect of SMS1 function that is being affected. Even if no further experiments were carried out, the authors could discuss possibilities. One might speculate what the PS is being used for. For example, is it a co-factor for integral membrane proteins, such as flippases? Is it a co-factor for peripheral membrane proteins, such as yet more LTPs? The model could include the work of Peretti et al (2008), which linked Nir2 activity exchanging PI:PA (Yadav et al, 2015) to the eventual function of CERT. Could the PS have a role in removing/reducing DAG produced by CERT?

Reviewer #3 (Public Review):

Summary:

The authors developed a lipid transfer protein knock-out library to identify lipid transfer proteins with roles in lipid homeostasis/metabolism. They investigated one of their hits, the ORP9/ORP11 dimer, which they found affects sphingomyelin synthesis. Further. they found that ORP9/11 localizes to ER-Golgi contact sites via interactions between a known FFAT motif in ORP9, which can interact with the ER protein VAP, and the PH domains of ORP9 and ORP11 that target PI4P at the Golgi. They showed defects in Golgi PI4P and PS levels when ORP9 or 11 were dysfunctional, supporting but not demonstrating that ORP9/11 might exchange PI4P and PS at these contact sites. Their in vitro data indicates that both ORP9 and 11 can transfer PS. They do not assess whether either protein can transfer PI4P (although there is a very nice recent paper by He et al et You, PMID 36853333, showing that ORP9 can transfer PI4P in vitro), and they do not assess the consequences of heterodimerization on either PS or PI4P transfer. The mechanisms by which PI4P/PS level perturbations affect sphingomyelin synthesis remain unclear.

Strengths:

The authors have developed an LTP knock-out library that might generate hypotheses regarding lipid metabolism, although defects are not unexpected and mechanisms will be difficult to work out--as, in fact, evidenced by this manuscript.

The OPR9/11 localization data and imaging studies are well done; this is the first more comprehensive characterization of the ORP9/11 heterodimer, which was discovered in 2010.

Weaknesses:

A major flaw is that the authors claim to but do not, in fact, provide evidence of PS/PI4P counter exchange in vitro. That the presence of PI4P on the acceptor liposomes accelerates PS transfer in the in vitro assays is not proof that there is a counter exchange. In fact, since the rate-determining step in the transfer reactions is lipid transfer between membrane and transfer protein and this depends on the association of the transfer protein with donor and/or acceptor liposomes, a more likely explanation for the more efficient transfer in the presence of PI4P is that PI4P allows for longer association of lipid transfer protein with acceptor liposomes. To show the plausibility of the counter-exchange idea as applied to the ORP9/ORP11 heterodimer, the authors would need to show that it can transfer PI4P. (The work by He et al et You, 2023, mentioned above, is a very nice study that the authors might use as a model.)

Mechanistic insights from the study are limited. How does a PI4P/PS imbalance affect sphingomyelin synthesis?

The ORP9/11 heterodimer seems to behave very much like ORP9/ORP10 heterodimer, including in its localization and dimerization mode. Is ORP9/11 just another version of 9/10? I wonder whether discussions of whether they are redundant or what their different roles are might be in order. There is little mechanistic or conceptual novelty arising from this study.

A minor point, but the statement (p2., lines 19-20) that "vesicular trafficking contributes only to a small portion of lipid trafficking" is not correct and raises eyebrows. What is more correct is that protein-mediated lipid transfer ALSO plays an important role in lipid transfer. It might even be said that LTP-mediated lipid transfer is critical in fine-tuning membrane lipid composition, including of phosphoinositides.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

The authors should possibly discuss more the other cases when LTPs of the same type of ORP9 and ORP10 have been found to dimerise. They should definitely cite and discuss the evidence reported in February this year in CMLS (see https://link.springer.com/article/10.1007/s00018-023-04728-5). In this paper, authors reported very similar findings as those the authors have in Figures 3, 4, S6, S7, and S8. Specifically, in this CMLS paper the authors find that ORP9 and ORP10 (not ORP11) interact through a central helical region and that ORP9 localises ORP10 to the ER-Golgi MCSs by providing ORP10 with a binding site for VAPs, where the heterodimer mediates the exchange of PtdIns(4)P for PtdSer.

We thank the reviewer for their recommendations. The mentioned paper has simply gone unnoticed by us and is now referred in the revised manuscript. Various other papers reporting on LTP dimerizations are already cited in our manuscript: ORP9-ORP10 dimerization (Kawasaki et al. 2022), ORP9-ORP11 dimerization (Zhou et al. 2010), and ORP9-ORP10/11 dimerization (Tan and Finkel 2022). Revised manuscript now discusses the dimerization of CERT and OSBP while citing Gehin et al. 2023, Ridgway et al. 1992 and de la Mora et al. 2021.

Reviewer #2 (Recommendations For The Authors):

Model and Discussion:

Give an idea about the aspect of SMS1 function that is being affected. Even if no further experiments were carried out, the authors could discuss possibilities. One might speculate what the PS is being used for. For example, is it a co-factor for integral membrane proteins, such as flippases? Is it a co-factor for peripheral membrane proteins, such as yet more LTPs? The model could include the work of Peretti et al (2008), which linked Nir2 activity exchanging PI:PA (Yadav et al, 2015) to the eventual function of CERT. Could the PS have a role in removing/reducing DAG produced by CERT?

We thank the reviewer for their recommendations. The same recommendations were also scripted in the public review, which we believe we answered sufficiently.

Other, Minor:

Make clear that there is no sterol readout (Fig 1C)

We would like to point out that Figure 1C has a sterol readout as CE refers to cholesterol esters.

PH domains of ORP9 and ORP11 localized only partially to the Golgi, unlike the PH domains of OSBP and CERT" (line 154). Say here where the non-Golgi ORP9 and ORP11 PH domain pool is - presumably in the cytoplasm.

We thank the reviewer for their suggestion and rephrase the sentence accordingly.

Fig 7H-J: histograms not lines as these are separate unlinked categories

We thank the reviewer for their suggestion. However, we think the original figure represent our findings in the best possible way. Our analysis regarding individual lipid species is also included in Supplementary figure 10.

Reviewer #3 (Recommendations For The Authors):

(1) At the end of the intro, in summarizing their findings, the authors state (p3. lines 48-49) "These findings highlight how phospholipid and sphingolipid gradients along the secretory pathway are linked at ER-Golgi membrane contact sites." This should instead read "These findings highlight THAT phospholipid and sphingolipid gradients along the secretory pathway are linked at ER-Golgi membrane contact sites."

We thank the reviewer for their suggestion and change the sentence accordingly.

(2) As noted in the public section, to show that ORP9/11 do indeed exchange lipids, an in vitro experiment demonstrating that ORP11 can transfer PI4P is essential. Ideally, it would be best to examine PS AND PI4P transfer by ORP9 AND 11 separately AND then by the ORP9/11 heterodimer. This could lend insights as to the function of the heterodimer. The He et al et Yu paper should provide guidelines for this. Why have the heterodimers?

We believe we addressed this point by showing the lipid transfer ability of the ORP9-ORP11 dimer. These findings are now part of the revised manuscript.

(3) It would be interesting to discuss the roles of ORP9/ORP11 versus ORP9/ORP10... they seem so analogous, although this is at the discretion of the authors.

We thank the reviewer for their suggestion. Since the difference between ORP9-ORP10 and ORP9-ORP11 dimers was also raised by other reviewers, we decided to include this discussion in the manuscript. A section based on our answer to Reviewer #2 in Public Review is now part of the Discussions.

(4) The authors used a melanoma cell line in their screens (p3, line 59). Could they explain why they used this cell line versus others?

We chose MelJuSo cell for various reasons. Mainly, MelJuSo are diploid, which eases generating knockouts in a screening setup compared to other polyploid cancer cell lines (e.g. HeLa). Furthermore, our CRISPR/Cas9 screening protocols are optimized for these cell lines.

Version 2

Abstract

Abstract

Numerous lipids are heterogeneously distributed among organelles. Most lipid trafficking between organelles is achieved by a group of lipid transfer proteins (LTPs) that carry lipids using their hydrophobic cavities. The human genome encodes many intracellular LTPs responsible for lipid trafficking and the function of many LTPs in defining cellular lipid levels and distributions is unclear. Here, we created a gene knockout library targeting 90 intracellular LTPs and performed whole-cell lipidomics analysis. This analysis confirmed known lipid disturbances and identified new ones caused by loss of LTPs. Among these, we found major sphingolipid imbalances in ORP9 and ORP11 knockout cells, two proteins of previously unknown function in sphingolipid metabolism. ORP9 and ORP11 form a heterodimer to localize at the ER-trans Golgi membrane contact sites, where the dimer exchanges phosphatidylserine (PS) for phosphatidylinositol-4-phosphate (PI(4)P) between the two organelles. Consequently, loss of either protein causes phospholipid imbalances in the Golgi apparatus that result in lowered sphingomyelin synthesis at this organelle. Overall, our LTP knockout library toolbox identifies various proteins in control of cellular lipid levels, including the ORP9-ORP11 heterodimer, which exchanges PS and PI(4)P at the ER-Golgi membrane contact site as a critical step in sphingomyelin synthesis in the Golgi apparatus.

Introduction

Many types of lipids are heterogeneously distributed among various organelles of the cell1. Due to the hydrophobic nature of lipids, their trafficking needs to be facilitated. While vesicular trafficking transports lipids in bulk between organelles, non-vesicular trafficking of lipids also play a role in defining lipid compositions of organelles. In fact, many organelles that do not receive any vesicular cargo rely on this route. Non-vesicular trafficking of lipids is achieved by a group of lipid transfer proteins (LTPs) that carry lipids using their hydrophobic cavities to stabilize lipids in the aqueous intracellular environment2. LTPs often contain targeting domains, motifs, transmembrane regions, amphipathic helices or surface charge to define their donor and acceptor organelles3. Many LTPs localize at two organelles simultaneously at membrane contact sites, where these organelles come closer to exchange information and material, including lipids4,5.

The human genome encodes over 150 LTPs; new LTPs and LTP families are being characterized every so often2,6–13. About 50 of these proteins are secreted and are involved in carrying lipids, metals, lipopolysaccharides, and other small molecules in blood plasma2. The remaining intracellular LTPs are mainly responsible for intracellular lipid trafficking among organelles. Since many lipid-modifying enzymes are localized to different organelles, LTPs also feed metabolic lipid fluxes, eventually defining cellular lipid levels14. For example, LTP-mediated trafficking of ceramide from the endoplasmic reticulum (ER) to the trans Golgi is needed for sphingomyelin synthesis15. While achieving a considerable understanding of some LTPs in the last years, the function of many LTPs in defining cellular levels and distributions remains unclear.

To study LTPs systematically, we designed and created an arrayed gene knockout library targeting 90 intracellular LTPs based on the CRISPR/Cas9 technology. Lipidomics analysis of the library-generated LTP knockout cells confirmed known and identified novel lipid disturbances emerging from loss of LTPs. These included CERT, GLTP, NPC1 and NPC2 knockout cells with altered sphingolipid levels. Furthermore, we identified major sphingolipid imbalances in ORP9 and ORP11 knockout cells, two proteins of unknown function in sphingolipid metabolism. ORP9 and ORP11 form a heterodimer to localize at the ER-trans Golgi membrane contact sites. At this contact site, the ORP9-ORP11 dimer transfers phosphatidylserine (PS) from the ER to the Golgi and phosphatidylinositol-4-phosphate (PI(4)P) in the opposite direction. Consequently, loss of either protein causes phospholipid imbalances in the Golgi apparatus that results in lowered sphingomyelin synthesis capacity at this organelle. Collectively, our LTP knockout library toolbox identified various proteins controlling cellular lipid levels. Among these, we found the ORP9-ORP11 heterodimer defining phospholipid composition of the trans Golgi as a critical step in sphingomyelin synthesis. These findings highlighted that phospholipid and sphingolipid gradients along the secretory pathway are linked at the ER-Golgi membrane contact sites.

Results

A CRISPR knockout library targeting lipid transfer proteins

To understand the function of LTPs in defining cellular levels and distributions, we created an arrayed CRISPR/Cas9 knockout library targeting LTPs. We targeted only the intracellular LTPs in the library and excluded extracellular proteins, such as LIPOCALIN and LBP-BPI-CETP families2. In this library, 90 wells of a 96-well plate were used for targeting LTPs and 6 for non-targeting (NT) controls (Fig. 1A). For optimal gene disruption, we used a lentiviral gene delivery system and 3 guide RNAs per gene – a method also used by others to increase the frequency of gene disruption16. We applied the LTP knockout library to MelJuSo, a human melanoma cell line. The efficiency of the knockout strategy was confirmed by western blotting of selected wells (Fig. 1B).

Lipidomics analysis of a LTP knockout library revealed known and candidate controllers of cellular lipid levels.

(A) Overview of the arrayed gene knockout library targeting LTPs. NT: non-targeting. (B) Western blotting of the selected wells showing the efficiency of the LTP knockout library on MelJuSo cells. gNT: Non-targeting, control cells. (C) Summary plot of the lipidomics analysis: Z-scores calculated for 17 lipid classes are plotted. Each data point represents the mean Z-score of an LTP knockout cell line; size and color of data points represent the standard deviation of Z-scores for 3 experiments on average. PC: phosphatidylcholine, LPC: lyso-phosphatidylcholine, PE: phosphatidylethanolamine, LPE: lyso-phosphatidylethanolamine, PS: phosphatidylserine, PI: phosphatidylinositol, PA: phosphatidic acid, PG: phosphatidylglycerol, DCer: dihydroxyceramide, Cer: ceramide, SM: sphingomyelin, HexCer: hexosylceramide, LacCer: lactosylceramide, DG: diacylglycerol, TG: triacylglycerol, CE: cholesterol ester, FFA: free fatty acids.

Next, we performed whole-cell lipidomics analysis of the LTP knockout cells. To this end, library-generated MelJuSo cells were grown in lipid-depleted serum and analysed using a lipidomics platform detecting 17 different lipid classes thrice on average (Fig. 1A). One of the detected lipid classes, hexosylceramide, can correspond to glucosylceramide and galactosylceramide – two isomeric lipids in mammalian cells. By metabolic chasing of a fluorescent ceramide analogue in cells silenced for glucosylceramide synthase or galactosylceramide synthase; or treated with the glucosylceramide synthase inhibitor PDMP, we validated that glucosylceramide is the major hexosylceramide in MelJuSo cells (Fig. S1). This further substantiated the previous reports of galactosylceramide being present mainly in oligodendrocytes17.

We next calculated the z-scores within each lipid class based on their relative abundance per measurement and plotted their average for LTP knockout cell lines. The analysis revealed many LTP knockout cells with lipid imbalances (Fig. 1C, Fig. S2A). Meanwhile, no z-score above the absolute value of 2 was observed for NT control cells (Fig. S2B). Overall, the lipidomics analysis of the LTP knockout library uncovered numerous candidate regulators of cellular lipid levels.

NPC1 and NPC2 knockout cells accumulate sphingomyelin in lysosomes and the plasma membrane

Lipidomics analysis of the LTP knockout library revealed various anticipated imbalances. Among these candidate regulators of cellular lipid levels, those related to sphingolipid metabolism are notable. This is likely due to many enzymatic steps of sphingolipid metabolism being more linear and lacking the elasticity of phospholipids1,14. In the analysis, we identified GLTP knockout cells with lowered glucosylceramide levels (Fig. 2A, Fig. S3A-C). Loss of GLTP is recently reported to block ER-to-Golgi anterograde vesicular trafficking that results in lowered glucosylceramide synthesis18.

NPC1 and NPC2 knockout cells accumulate sphingomyelin in lysosomes.

(A) GLTP knockouts show decreased glucosylceramide levels compared to control cells. GlcCer: glucosylceramide (B) NPC1 and NPC2 knockout cells have increased sphingolipids levels compared to control cells. (C) Increased sphingolipid levels in NPC1 and NPC2 knockouts are represented in all sphingolipid classes. (D) Immunofluorescence images of NPC1 and NPC2 knockout cells permeabilized and stained with recombinant GFP-EQT show the accumulation of sphingomyelin in the lysosomes compared to the parental MelJuSo (MJS) cells. Pixel intensities of the images were adjusted evenly. Scale bar 10µm. (E) Flow cytometry analysis of cells stained with EQT-GFP, CTB-Alexa568 and CD63-V450 on the cell surface. Analysis demonstrates the accumulation of sphingomyelin (EQT-GFP) and the glycosphingolipid GM1 (CTB-Alexa568) on the cell surface without affecting the surface protein levels. (F) Mean fluorescence intensities for each staining normalized to control cells from 3 experiments. Red lines and error bars correspond to mean and standard deviations from 3 experiments, respectively. Approximately 2000 cells were analyzed per experiment.

In addition, we observed NPC1 and NPC2 knockout cells with elevated sphingolipid levels (Fig. 1B). Mutations in NPC1 and NPC2 genes are the cause of Niemann-Pick disease, type C that leads to lysosomal accumulation of cholesterol19. Cells lacking NPC1 or NPC2 function also accumulate sphingolipids20. Our observation of elevated sphingolipid levels in NPC1 and NPC2 knockout cells was present in all sphingolipid classes (Fig. 2B, C). Meanwhile phospholipid levels were not altered dramatically in these cells (Fig. S3D). To confirm whether this increase is associated with lysosomal accumulation of sphingolipids, we used a sphingomyelin-binding biosensor based on the equinatoxin secreted by the beadlet anemone Actinia equina (Fig. S3E)21,22. Immunofluorescence staining of NPC1 and NPC2 knockout cells using the biosensor showed lysosomal accumulation of sphingomyelin (Fig. 2D). Flow cytometry analysis using the sphingomyelin biosensor and the GM1 glycosphingolipid biosensor, the B subunit of cholera toxin, demonstrated that sphingolipid accumulation is also reflected on the cell surface of NPC1 and NPC2 knockouts (Fig. 2E, F). Meanwhile, the surface protein levels were not altered as detected by staining for the tetraspanin protein CD63. Collectively, our results corroborate the sphingolipid accumulation in NPC1 and NPC2 deficient cells and illustrate that the LTP knockout library is a viable tool to study LTPs and their role in regulating cellular lipid levels.

CERT, ORP9 and ORP11 knockout cells demonstrate reduced sphingomyelin levels

Sphingomyelin is the major sphingolipid in mammalian cells, and it is mainly synthesized in the trans Golgi from ceramide delivered from the ER. The non-vesicular trafficking of ceramide from the ER to the trans Golgi is mediated by the ceramide transfer protein CERT15,23 (Fig. 3B). Our analysis identified CERT knockouts with the lowest sphingomyelin levels (Fig. 1C, 3A). These cells also demonstrate increased glucosylceramide levels, likely as a compensation mechanism (Fig. 3A). Furthermore, we found decreased sphingomyelin and increased glucosylceramide levels in ORP9 and ORP11 knockout cells – two proteins of unknown role in sphingolipid metabolism (Fig. 1C, 3A). The sphingolipid imbalances in all three knockout cells –CERT, ORP9, and ORP11– were also represented in many sphingolipid subspecies (Fig. 3C, Fig. S4B). Lowered sphingomyelin levels of the three knockout cells were further confirmed by their sensitivity towards methyl-β-cyclodextrin treatment (Fig. 3D). Meanwhile, none of the three knockout cells demonstrated phospholipid imbalances to a compelling degree (Fig. S4A).

CERT, ORP9 and ORP11 knockout cells demonstrate reduced sphingomyelin levels.

(A) CERT, ORP9 and ORP11 knockouts demonstrate decreased sphingomyelin and increased glucosylceramide levels. (B) Schematic representation of de novo sphingomyelin synthesis pathway of mammalian cells. (C) Decreased sphingomyelin and increased glucosylceramide levels observed in knockout cells are present in various sphingolipid subspecies. (D) Similar to CERT knockouts, ORP9 and ORP11 knockouts are sensitive to methyl-β-cyclodextrin treatment. Red lines correspond to mean values from 4 experiments. (E) Western blot of knockout cells showing that the loss of ORP9 or ORP11 does not affect CERT protein levels. (F) ORP9 and ORP11 localize at the Golgi apparatus. Scale bar 10µm. (G) Domain architecture of CERT, ORP9 and ORP11. All proteins contain PH domains for Golgi localization. CERT and ORP9, but not ORP11, contain a FFAT motif for interacting with VAP proteins.

The observation that ORP9 and ORP11 knockout cells mimic CERT knockouts implied a role for ORP9 and ORP11 in de novo sphingomyelin synthesis, whereas loss of neither ORP9 nor ORP11 reduced CERT protein levels (Fig. 3E). ORP9 and ORP11 each contain a PH domain mediating localization at the trans Golgi, the site of de novo sphingomyelin synthesis (Fig.3F, G). As loss of either ORP9 or ORP11 did not alter the Golgi ultrastructure as detected by light or electron microscopy (Fig. S5), it further suggested that ORP9 and ORP11 affect sphingomyelin levels by a previously unknown mechanism. Next, we investigated the possible mechanism by which ORP9 and ORP11 regulate cellular sphingolipid levels.

Dimerization is critical for the ER-Golgi contact site localization of ORP9 and ORP11

ORP9 and ORP11 interact with each other via their regions between their PH and ORD domains24. As no other domain is found in this region, we speculated some secondary structures must facilitate the dimerization. AlphaFold and PCOILS coiled coils prediction tool suggested that ORP9 and ORP11 contain two alpha helices each in this region – hereafter referred as “coils” (Fig. 4A, S6A, B). AlphaFold-Multimer suggested that the coils of ORP9 and ORP11 interact with each other (Fig. 4B, S6C-D). Identical analysis of the full-length proteins predicted that the coils drive dimerization of the two proteins (Fig. 4C, D, S6E). This was tested by targeting the coils of ORP9 to the mitochondrial outer membrane using the N-terminal sequence of TOM70. In cells expressing mitochondria-targeted ORP9 coils, the coils of ORP11 located to this organelle (Fig. 4C, D). The absence of either coil prevented the colocalization, suggesting that the coils of ORP9 and ORP11 are sufficient for their dimerization. This finding was validated by co- immunoprecipitations (Fig. 4E).

ORP9 dimerization is critical for ER localization of ORP11.

(A) AlphaFold revealing coiled coils in ORP9 and ORP11. (B) AlphaFold-Multimer shows that coiled coils of ORP9 and ORP11 interacting with each other. (C) ColabFold protein complex prediction for full length ORP9 and ORP11 shows the dimerization via coiled coils (arrow). (D) Position aligned error for panel C. Note that the interaction of coils has a lower score (yellow arrows). (E) Coiled coils of ORP9 and ORP11 is sufficient to define their localization as ORP11 coiled coils colocalizes with the mitochondria-targeted ORP9 coiled coils at mitochondria. Scale bar: 10µm. (F) Colocalization analysis of immunofluorescent images corresponding to panel E. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (G) Co- immunoprecipitation confirming that the coils of ORP9 and ORP11 interact with each other. (H) Immunofluorescence images of PH domain localizations. Compared to individual PH domains of ORP9 and ORP11, ORP9-ORP11 chimera demonstrates better affinity towards the Golgi. Scale bar: 10µm. (I) Colocalization analysis of immunofluorescent images corresponding to panel H. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (J) Co-immunoprecipitation analysis of VAPA, ORP9 and ORP11 from over-expressing cells. Despite lacking a FFAT motif, ORP11 interacts with the ER-resident VAPA protein. This interaction is facilitated by the FFAT motif of ORP9 as a VAPA mutant unable to interact with FFAT motifs (VAPA**) was unable to co-precipitate ORP9 or ORP11.

ORP9, but not ORP11, contains a FFAT, two phenylalanines in an acidic tract, motif that drives ER localization by interacting with the ER-resident VAPA and VAPB25. Despite not containing a FFAT motif, ORP11 could interact with VAPA and VAPB by its dimerization with ORP9. We tested this by co- immunoprecipitations, where ORP11 could be co-isolated with VAPA from cells expressing VAPA, ORP9 and ORP11 (Fig. 4J). This interaction was diminished significantly when ORP9 was not co-expressed. Furthermore, a VAPA mutant unable to interact with FFAT motifs failed to interact with either ORP9 or ORP11, showing that ORP11 interacts with VAPA indirectly via the ORP9-FFAT motif.

We and others previously showed that in addition to VAPA and VAPB, the human proteome contains three other proteins interacting with FFAT and related short linear motifs: MOSPD1, MOSPD2 and MOSPD326,27. In our earlier efforts to identify interaction partners of VAPA, VAPB and other motif-binding proteins, we performed BioID, proximity biotinylation followed by proteomics, identification of membrane contact sites proteins27. In this analysis, ORP9 and ORP11 were found in proximity with VAPA and VAPB but not with other motif-binding proteins (Fig. S6F). This further confirmed that despite lacking a FFAT motif, ORP11 together with ORP9 is part of VAPA- and VAPB-mediated membrane contact sites.

The Golgi localization of ORP9 and ORP11 is mediated by their PH domain interacting with phosphatidylinositol phosphates24. We observed that the PH domains of ORP9 and ORP11 localized only partially to the Golgi and demonstrated strong cytoplasmic localization, unlike the PH domains of OSBP and CERT that show exclusive localization to this organelle28,29 (Fig. 4H, S7D). We hypothesized that the dimerization of ORP9 and ORP11 is a mechanism to increase their avidity towards trans Golgi membranes. To test this, we created a chimera containing ORP9- and ORP11-PH domains that localized to the Golgi more efficiently than the individual PH domains (Fig. 4H, I). This also implied that loss of either protein should reduce the other’s localization to the trans Golgi. Immunofluorescence staining of ORP9 and ORP11 knockout cells for the same proteins confirmed that both ORP9 and ORP11 are required for the Golgi localization of the other protein (Fig. 5). Meanwhile, loss of either protein did not influence the levels of the other (Fig. 3E). Our results collectively show that the dimerization of ORP9 and ORP11 via their coils is required for the localization of ORP9-ORP11 dimer to the ER and the trans Golgi simultaneously.

ORP9 and ORP11 dimerization is critical for their Golgi localization.

(A) Immunofluorescence images of cells stained for ORP9. ORP9 fails to localize at the Golgi as efficiently in ORP11 knockout cells. (B) Colocalization analysis of immunofluorescent images corresponding to panel A. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (C) Immunofluorescence images of cells stained using an ORP11 antibody. Similar to ORP9, ORP11 fails to localize at the Golgi as efficiently in ORP9 knockout cells. Note that the effect of ORP9 loss on ORP11 localization is more dramatic than vice versa. (D) Colocalization analysis of immunofluorescent images from panel C. Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. All scale bars are 10µm.

ORP9 and ORP11 are essential for PS and PI(4)P levels in the Golgi apparatus

ORP9 and ORP11 belong to the OSBP-related protein (ORP) family that transfers sterols or PS while transferring phosphatidylinositol phosphates in the opposite direction30–33 (Fig. 6A, S7B). ORP9 and ORP11 carry the conserved PS binding site and are recently shown to relocate to the site of lysosomal damage to supply PS for membrane repair34. Using an in vitro lipid transfer assay, we confirmed that the lipid transfer domains (ORD) of ORP9 and ORP11 are sufficient to traffic PS in vitro (Fig. 6B, C). ORP9 and ORP10, but not ORP11, are shown to traffic PI(4)P35,36. Using a PI(4)P transfer assay that utilizes the PI(4)P-binding PH domain of FAPP1, we show that ORD of ORP11 can also traffic PI(4)P between model membranes (Fig. 6D, E, S7C, D)37. Furthermore, similar to ORP5, ORP8 and ORP10, the PS-trafficking activity of ORP9- and ORP11-ORDs was improved when acceptor liposomes were decorated with PI(4)P35,38,39 (Fig. S7E). This indicated that ORDs of ORP9 and ORP11 transport PI(4)P in the opposite direction of PS trafficking.

ORP9 and ORP11 are essential for PS and PI(4)P levels in the Golgi apparatus.

(A) Phylogenetic tree of human OSBP-related domains showing that ORP9 and ORP11 belong to the PS transporter branch. (B) Schematic representation of FRET-based PS transfer assay. Rhodamine in the donor vesicles quenches NBD fluorescence unless NBD-labeled lipid is transferred to acceptor liposomes. (C) FRET- based lipid transfer assay using the OSBP-related domains of ORP9 and ORP11, ORD9 and ORD11 respectively, demonstrate the PS trafficking ability between vesicles in vitro. (D) Graphical representation of PI(4)P transfer assay that utilizes the NBD-labelled PH domain of FAPP1. PI(4)P transfer by a protein leads to inrease NBD-fluorescence by dequenching. (E) PI(4)P transfer assay using ORP9-ORD and ORP11-ORD (ORD9 and ORD11, respectively). (F) Schematic representation of in vitro reconstitution of membrane contact sites for lipid transfer assays. (G) PS transfer assay at in vitro reconstituted membrane contact sites formed by the VAPB-ORP9-ORP11 trimer. (H) Vmax quantifications of PS transfer assays from panel E. (I) The PS sensor Lact-C2 localizes less prominently to the Golgi in ORP9 and ORP11 knockout cells. (J) Golgi quantification of the PS sensor RFP-LactC2 indicates reduced PS levels in this organelle. This phenotype could be rescued by reconstitution of the missing protein in knockout cells (shown with a dagger). Red lines correspond to mean values from 3 experiments; n is number of analyzed cells. (K) The PI(4)P sensor OSBP- PH localizes more prominently to the Golgi area in ORP9 and ORP11 knockout cells. (I) Golgi quantification of the PI(4)P sensor confirms the accumulation of PI(4)P at this organelle. This phenotype was rescued by reconstitution of the missing protein in knockout cells (shown with a dagger). Red lines correspond to mean values from 3 experiments; n is number of analyzed cells.

To demonstrate the contribution of ORP9-ORP11 dimerization to their lipid transfer ability at membrane contact sites, we reconstituted these sites in vitro by attaching a His-tagged VAPB devoid of its transmembrane helix to donor liposomes and providing full-length ORP9 and ORP11 as tethers to the acceptor liposomes (Fig. 6F, S7F-H). PS transfer assay in these reconstituted contact sites showed that as a trimer, the ORP9-ORP11-VAPB complex displays an improved lipid transfer capacity (Fig. 6G, H).

PS is synthesized in the ER and its concentration increases along the secretory pathway as it is enriched on the cytosolic leaflet of the plasma membrane14. PI(4)P, on the other hand, is abundant in the trans Golgi, where it recruits many proteins to this membrane40. Another LTP localized at the ER-trans Golgi membrane contact sites, OSBP transfers cholesterol from the ER to the trans Golgi, while counter transporting PI(4)P to the ER for its hydrolysis by the ER-resident PI(4)P phosphatase SAC1 (Fig. 8). Homologous to OSBP, the ORP9-ORP11 dimer resides in the ideal intracellular interface to traffic PS in anterograde and PI(4)P in retrograde direction between the ER and trans Golgi (Fig. 8). This also suggested that the loss of either protein would cause PS and PI(4)P imbalances between the ER and trans Golgi. We tested this notion by using PI(4)P- and PS-binding biosensors41,42 (Fig. S7I, J). Golgi localization of these sensors indicated that PS levels were decreased in the Golgi of these cells (Fig. 6I, J). Also, PI(4)P showed accumulation in the Golgi of ORP9 and ORP11 knockout cells (Fig. 6K, L). We confirmed the PI(4)P accumulation by quantifying the Golgi localization of a PI(4)P specific antibody (Fig. S7J). Interestingly, both phenotypes, lowered PS and increased PI(4)P levels, were better pronounced in ORP9 knockout cells compared to ORP11 knockouts. Furthermore, both phenotypes could be rescued by reconstitution of the missing protein (Fig. 6J, L). Overall, these findings suggested that the ORP9-ORP11 dimer is required for maintaining PS and reducing PI(4)P levels in the Golgi apparatus.

De novo sphingomyelin synthesis in the Golgi is impaired in ORP9 and ORP11 knockouts

Next, we set out to investigate the possible role of ORP9 and ORP11 in regulating cellular sphingomyelin levels. By performing a sphingomyelin synthase activity assay that uses the fluorescent ceramide analogue NBD-ceramide as a substrate43, we found that ORP9 and ORP11 knockout cells do not have reduced sphingomyelin synthesis capacity (Fig. 7A, B). The same was also found in CERT knockout cells. Whereas all three knockouts demonstrated increased glucosylceramide synthesis capacity, supporting the lipidomics analysis (Fig. 3A). Unaltered sphingomyelin synthesis capacity was further validated by the unreduced protein and mRNA levels of both human sphingomyelin synthases, SMS1 and SMS2 (Fig. S8A, B).

ORP9 and ORP11 are needed for de novo sphingomyelin synthesis in the Golgi apparatus.

(A) Thin-layer chromatography readout of the enzymatic activity assay performed in lysates reveals the unreduced sphingomyelin synthesis capacity of knockout cells. (B) Quantification of the enzymatic activity assays corresponding to panel A. Note the increased GlcCer production capacity in the knockout cells. Red lines and error bars correspond to mean and standard deviation from 5 experiments, respectively. (C) Graphic representation of experiments using palmitic acid alkyne for monitoring de novo sphingomyelin synthesis. Alkyne modified sphingolipids are “clicked” with azido-coumarin before thin-layer chromatography (TLC) analysis. (D) TLC readout of de novo sphingomyelin synthesis assay in intact cells using palmitic acid alkyne. Knockout cells demonstrate reduced conversion of palmitic acid to sphingomyelin. FA: fatty acid, SM: sphingomyelin, LPE: lyso-O-phosphatidylethanolamine (E) Quantification of de novo sphingomyelin synthesis assay from panel C. Red lines and error bars correspond to mean and standard deviation from 5 experiments, respectively. (F) Simplified representation of Golgi-targeted lipidomics and enzymatic activity assays. (G) Western blot confirming the immunomagnetic isolation of trans Golgi membranes. Cross-reactivity to Protein G and/or Golgin-97 IgG were labeled with asterisks. (H-J) Lipidomics analysis of Golgi isolates show lowered PS levels in the Golgi of ORP9 and ORP11 knockout cells. Same Golgi fractions have reduced sphingomyelin but increased ceramide levels. Each colored line corresponds to a lipid species. Data points and the error bars correspond to mean and standard deviations from 4 experiments. (K) TLC readout of enzymatic activity assay performed in isolated Golgi from knockout cells. Golgi of ORP11 and especially that of ORP9 demonstrate lowered capacity to synthesize sphingomyelin. (L) Quantification of sphingomyelin synthesis activity assay performed in Golgi isolates corresponding to panel I.

Model of ORP9-ORP11 mediated promotion of sphingomyelin synthesis at the ER-trans Golgi contact site.

ORP9 and ORP11 require dimerization with each other to localize at ER-trans Golgi membrane contact sites, where they exchange PS for PI(4)P. Consequently loss of either protein causes PS and PI(4)P imbalances in the Golgi apparatus. Since only ORP9 contains a FFAT motif, its loss leads to a more pronounced effect on PS, PI(4)P and sphingolipid levels. The same contact site accommodates other LTPs. Multiple phosphatidylinositol transfer proteins (PITPs) traffic phosphatidylinositol to the Golgi for their phosphorylation. Each PITP at the ER-Golgi contact site serves a different function—one of these proteins, Nir2, also contains a FFAT motif (ref.46 and 54). OSBP and CERT are responsible for the anterograde trafficking of cholesterol and ceramide, the latter is used for sphingomyelin production. Cholesterol, sphingomyelin and PS are trafficked to the plasma membrane by vesicular means to be asymmetrically distributed between the leaflets of the plasma membrane.

SMS1 is localized at the trans Golgi, while SMS2 mainly localizes at the plasma membrane but is also found at the trans Golgi44. As the enzymatic activity assay in lysates reports on the global sphingomyelin synthesis capacity, we next investigated de novo sphingomyelin synthesis that occurs at the trans Golgi, where CERT as well as the ORP9-ORP11 dimer localize. For this purpose, we chased the metabolic fate of palmitic acid alkyne (Fig. 7C). Palmitic acid is the precursor of sphinganine and all sphingolipids (Fig. 3B, 7C). To enter the sphingolipid pathway, palmitic acid first needs to travel to the ER and later to the Golgi for conversion to sphingomyelin, unlike ceramide analogues, such as NBD-ceramide, that can be converted to sphingomyelin by the plasma membrane-resident SMS2. Furthermore, the palmitic acid alkyne does not contain a bulky fluorescent group and the terminal alkyne allows visualization by click chemistry (Fig. 7C). As this method enabled us to monitor de novo sphingomyelin production in intact cells, we confirmed the reduced de novo sphingomyelin synthesis in CERT knockout cells using this method (Fig. 7D, E, S8C). In addition, ORP9 and ORP11 knockouts also showed decreased conversion of palmitic acid to sphingomyelin, suggesting a similar defect as CERT knockouts. To confirm the decreased sphingomyelin synthesis, we quantified the Golgi localization of a DAG biosensor, C1ab domain of protein kinase D1, that can report on sphingomyelin synthesis45,46. This showed decreased DAG levels in the Golgi that can be accounted for lowered sphingomyelin synthesis (Fig. S8D).

The observations that loss of ORP9 or ORP11 does not affect CERT protein (Fig. 3E) or reduce CERT localization to the Golgi (Fig. S9) implied that CERT-mediated transfer routes are not affected in ORP9 or ORP11 knockouts. Lipidomics analysis showing accumulation of ceramide and glucosylceramide in ORP9/ORP11 knockouts further suggested the conversion of ER-bound ceramides to sphingomyelin in the Golgi is hampered.

To distinguish between two possibilities that may result in lowered sphingomyelin levels – a ceramide delivery defect to the trans Golgi or an inability to convert ceramide to sphingomyelin in the Golgi, we established an immunoisolation protocol for trans Golgi membranes using a monoclonal antibody against the trans Golgi marker Golgin-97 (Fig. 7F). These isolates were enriched for the trans Golgi marker Golgin- 97 and were devoid of other organelle markers, such as the ER, mitochondria, and lysosomes (Fig. 7G). We confirmed the membrane integrity of Golgin-97-enriched fractions by their sphingomyelin synthase activity, as these multi-pass transmembrane proteins require intact membranes for activity44 (Fig. S10A). Lipidomics analysis of these fractions validated that the loss of ORP11 and especially that of ORP9 results in lowered PS levels in the trans Golgi (Fig. 7H, S10B-D). The trans Golgi fractions also showed reduced sphingomyelin levels, confirming the limited de novo sphingomyelin synthesis (Fig. 7D, E, I S10E). Moreover, the same fractions showed elevated ceramide levels in these fractions (Fig. 7J, S10F), revealing that the knockouts do not have a ceramide delivery defect, but instead a lowered capacity to convert ceramide to sphingomyelin in their trans Golgi membranes. We tested this notion by performing a sphingomyelin synthase activity assay in the isolated Golgi membranes, where supplying excess substrate to these membranes bypasses the ceramide delivery routes (Fig. 7K, L). This assay further substantiated that the Golgi of ORP11 and especially of ORP9 knockout cells have a lowered sphingomyelin synthesis capacity. Meanwhile, localization of both sphingomyelin synthases was not reduced in the Golgi, as detected by three different antibodies (Fig. S11). Collectively, our results indicate that PS-PI(4)P exchange between the ER and trans Golgi mediated by the ORP9-ORP11 LTP complex is critical for de novo sphingomyelin synthesis in trans Golgi.

Discussion

In this study, we describe a gene knockout library for systematic characterization of intracellular LTPs. The arrayed design of the library enables high and low throughput analysis. As new LTPs are identified in regular basis, the arrayed feature of the library allows expansion to include newly identified LTPs, e.g. ATG2A, SHIP164, KIAA0100 (Hobbit in Drosophila), KIAA1109 (Csf1 in yeast) and RMDN3 are new human LTPs that are identified through the course of this study8–10,12,13.

Lipidomics analysis of the library demonstrated many lipid imbalances. We further validated loss of NPC1, NPC2, CERT, and GLTP with sphingolipid imbalances15,18,20 (Fig. 1C). In addition to sphingolipids imbalances, we found STARD7 knockout cells with increased levels of phosphatidylglycerol (PG), a lipid class exclusive to mitochondria in mammals (Fig. 1C). STARD7 is a phosphatidylcholine (PC) transfer protein that localizes to the mitochondrial intermembrane space to supply PC to the mitochondrial inner membrane47–49. Consequently, loss of STARD7 reduces PC levels in this membrane as well as decreasing cardiolipin levels and respiratory capacity48,49. While PC in the inner membrane is not a precursor of cardiolipin, it is possible the reduced PC levels in this membrane impair the activity of the cardiolipin synthase that uses PG as its substrate, suggesting an explanation for the PG accumulation in STARD7 knockout cells.

LTPs often localize at membrane contact sites to facilitate lipid exchanges between organelles. We found ORP9 and ORP11 localizing at the ER-trans Golgi contact sites to exchange PS and PI(4)P. We show that the dimerization is critical for the contact site localization, and consequently, loss of either protein is sufficient to disturb PS and PI(4)P levels in the Golgi. In return, this phospholipid imbalance reduces sphingomyelin synthesis capacity of the Golgi. Compared to ORP11 knockout cells, ORP9 knockout cells display more pronounced phenotypes of cellular sphingolipid imbalances (sphingomyelin and glucosylceramide), phospholipid imbalances in the Golgi apparatus (PS and PI(4)P), as well as reduced capacity of sphingomyelin synthesis in trans Golgi membranes. Furthermore, loss of ORP9 leads to a more dramatic effect on the Golgi localization of ORP11 than vice versa. This is most likely due to ORP9 providing a critical factor, i.e. the FFAT motif, for contact site localization, whereas it is possible the loss of ORP11 can be partially compensated by ORP10.

The contact site localization of the ORP9-ORP11 dimer is homologous to that of the ORP9-ORP10 dimer localizing at the ER-endosome and ER-Golgi contact sites asORP10 also uses its coiled coils to interact with ORP935,36. In both LTP complexes, ORP9 provides the FFAT motif required for ER contact. ORP10 can also deliver PS to Golgi membranes50, however, our lipidomics analysis of ORP10 knockout cells did not demonstrate any sphingolipid imbalances (Fig. S12A). Besides ORP9, ORP10 and ORP11, other LTPs are also shown to dimerize at the ER-Golgi interface. A recent structural study has revealed the architecture of a previously-described OSBP dimerization that is mediated by a central core domain containing two alpha helices51,52. CERT forms a similar dimer and the conformational changes dictated by the central core domain is responsible for regulating the activity of the CERT dimer53. Both OSBP and CERT homodimers form a T-shaped structure, dissimilar to the heterodimers of ORP9, ORP10 and ORP11.

What distinguishes ORP11 from ORP10? An analysis of expression profiles from >900 cancer cell types showed that ORP10 and ORP11 are not differentially expressed (Fig. S12B). On the contrary, they show a positive correlation. It is possible that ORP9 interacts with ORP10 or ORP11 in different biological conditions or sub-cellular locations. ER-endosome contact site localization of the ORP9-ORP10 dimer supports this notion35. Even within the Golgi, ORP9-ORP10 and ORP9-ORP11 dimers might be facilitating lipid flows at different sub-organellar domains. The presence of multiple phosphatidylinositol (PI) transfer proteins localized at ER-trans Golgi contact sites feeding PI(4)P synthesis in the Golgi for distinct purposes and the observation of different PI(4)P-containing sub-domains in the Golgi promotes the idea that ORPs could localize at different Golgi sub-domains55,56. Furthermore, all three proteins, together with OSBP and ORP1L, are recruited to the lysosomal damage site in response to the increased PI(4)P levels on lysosomal delimiting membrane34,57. Collectively, these reports highlight the spatiotemporally dynamic nature of PI(4)P-mediated lipid trafficking and suggest that the two dimers could have different functions.

An intriguing observation is that the loss of ORP9 causes more accumulation of ceramide than CERT knockouts, despite loss of either protein leading to a comparable reduction of sphingomyelin levels. As the loss of CERT would lead to a ceramide accumulation in the ER, it is possible that such an accumulation in the ER is sensed to reduce ceramide production in this organelle. An ER-resident candidate ceramide sensor for this purpose was suggested previously58. Meanwhile, our findings show a ceramide accumulation in the trans Golgi of ORP9 knockouts. Accordingly, ceramide accumulation in the Golgi caused by the loss of ORP9 would fail to “turn on” an ER-localized ceramide sensing machinery, thus leading to further ceramide accumulation compared to CERT knockout cells. This notion also supports our finding that the loss of the ORP9-ORP11 dimer causes ceramide accumulation primarily in the Golgi.

Asymmetric distribution of lipids between two bilayer leaflets is a characteristic feature of the plasma membrane. This is owed to the build-up of sphingomyelin and cholesterol on the outer leaflet and PS on the inner leaflet14. The transition of the thin, symmetrical ER membrane to a thicker, rigid and asymmetrical one takes place in the Golgi (Fig. 8)14. Sphingolipid, cholesterol and PS concentrations also increase along the secretory pathway. Various mechanisms are described to drive the sphingolipid and cholesterol trafficking against the concentration gradient, including thermodynamic trapping of cholesterol due to complex formation with sphingomyelin or energy release from the retrograde trafficking of PI(4)P followed by its hydrolysis in the ER14,31,59. The same PI(4)P gradient between the ER and Golgi could also power the anterograde trafficking of PS, which, unlike sphingomyelin or cholesterol, is still exposed on the cytosolic leaflet of the Golgi where it has a higher concentration than the ER – thus, maintaining lipid trafficking against the gradient requires energy. In brief, our finding that the ORP9-ORP11-dimer-mediated phospholipid exchange promoting the sphingomyelin synthesis reveal further intertwining of lipid gradients along the secretory pathway at the ER-trans Golgi membrane contact sites.

Conflict of Interest

Authors declare no conflict of interest.